Data Munging

### 2020-10-05 scripts/dataMunge.Rmd
Added code to calculate upper mixed layer DO and temp.

---
title: "Data Munging"
authors: "J. Beaulieu, R. Martin, and M. McManus"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    fig_caption: yes
    depth: 2
    number_sections: true
editor_options: 
  chunk_output_type: console
---
```{r results='hide', message=FALSE, warning=FALSE, echo=FALSE}
# load libraries
library(plyr) # for ddply, load before tidyverse
library(tidyverse) # dplyr, ggplot
library(forcats) # fct_explicit_na()
library(rLakeAnalyzer) # for buoyancy frequency

@@

## Manipulate
Impliment unit conversion and create source/sink column:
``` {r}
dg <- dg %>%
  # unit conversion
  mutate(dissolved.ch4 = dissolved.ch4 * 1000000, # mol/L -> umol/L
         dissolved.co2 = dissolved.co2 * 1000000, # mol/L -> umol/L,
         dissolved.n2o = dissolved.n2o * 1000000000, # mol/L -> nmol/L
         # add source/sink column
         co2.src.snk = ifelse(co2.sat.ratio > 1, "source", "sink"),
         ch4.src.snk = ifelse(ch4.sat.ratio > 1, "source", "sink"),
         n2o.src.snk = ifelse(n2o.sat.ratio > 1, "source", "sink"))
```

@@

# Inspect results of join.  Looks good.  
# Only one site missing temperature and DO data. (NLA17_CA-10082)
# One site missing all DO data (NLA17_ID-10084)
temp.o21 %>% filter(SITE_ID %in% missing$SITE_ID) %>%
  select(SITE_ID, VISIT_NO, DEPTH, TEMPERATURE, OXYGEN) %>% 
  arrange(SITE_ID, VISIT_NO, DEPTH, TEMPERATURE, OXYGEN)

@@

@@

bf <- temp.o2.2 %>%
  filter(is.finite(TEMPERATURE.M)) %>% # This completely removes the CA site with only NA for temp data.
  group_split(SITE_ID, VISIT_NO) %>% # split into list of df
  lapply(function(x) {
    data.frame(BF = buoyancy.freq(wtr = x$TEMPERATURE.M, depths = x$DEPTH), # calculate buoyance frequency
               SITE_ID = unique(x$SITE_ID), # add site ID
               VISIT_NO = unique(x$VISIT_NO), # add visit no
               stringsAsFactors = FALSE)}) %>% 
  lapply(function(x) x %>% mutate(DEPTHS.BF = attr(BF, "depths"))) %>% # depths stored as attribute, assign to column
  do.call("rbind", .) # collapse to df

# there are three non-finite bf values, lets have a look
bf.na <- bf %>% filter(!is.finite(BF)) %>%
  select(SITE_ID, VISIT_NO) %>%
  mutate(SITE.VISIT = paste0(SITE_ID, VISIT_NO))

@@

# These site have a single temp measurement made at shallow depths (<1m).  Presumably these are 
# shallow sites, therefore unlikely to be stratified.  Assign a bf very close to 0 (see below)
bf %>% filter(paste0(SITE_ID, VISIT_NO) %in% bf.na$SITE.VISIT) 
temp.o2.2 %>% filter(paste0(SITE_ID, VISIT_NO) %in% bf.na$SITE.VISIT)

@@

`r bf.max %>% summarize(n.na = sum(!is.finite(bf.max$MAX.BF))) %>% pull()` of the site have a non-finite value for maximum buoyancy frequency value.  These are the three shallow sites with only a single temperature measurement.  Assign these the minimum bf.max value.
``` {r}
bf.max %>% filter(!is.finite(MAX.BF))

min.max.bf <- bf.max %>% 
  filter(is.finite(MAX.BF)) %>%
  summarize(min.max.bf = min(MAX.BF))

@@

@@

These calculations produced 8 site visits with at least one NA for surface/bottom temp/DO.

o2.surf at NLA17_FL-10029 is a case with no DO at shallowest depth, but good measurements below.  Replace with next shallowest value.
``` {r}
# Inspect raw data
temp.o2.2 %>% filter(SITE_ID == "NLA17_FL-10029", VISIT_NO == 1)

# update data
o2.s.b <- o2.s.b %>%
  mutate(o2.surf = replace(o2.surf, 
                           SITE_ID == "NLA17_FL-10029" & VISIT_NO == 1,
                           temp.o2 %>% filter(SITE_ID=="NLA17_FL-10029",
                                              VISIT_NO == 1,
                                              DEPTH == 1) %>%
                             select(OXYGEN) %>% pull()))

```
No bottom water DO or temp at NLA17_ID-10001, visit 1 due to probe malfunction.

```{r}
temp.o2.2 %>% filter(SITE_ID == "NLA17_ID-10001", VISIT_NO == 1)
```

No DO NLA17_ID-10084, visit 1 due to probe malfunction.

```{r}
temp.o2.2 %>% filter(SITE_ID == "NLA17_ID-10084", VISIT_NO == 1)
```

No deep DO or temp at NLA17_MI-10073, visit 1 because no measurement taken at deepest recorded depth.  Replace with next shallowest measurement.

```{r}
# look at raw data
temp.o2.2 %>% filter(SITE_ID == "NLA17_MI-10073", VISIT_NO == 1)

# update aggregated data
o2.s.b <- o2.s.b %>%
  mutate(o2.bottom = replace(o2.bottom, 
                           SITE_ID == "NLA17_MI-10073" & VISIT_NO == 1,
                           temp.o2.2 %>% filter(SITE_ID=="NLA17_MI-10073",
                                              VISIT_NO == 1,
                                              DEPTH == 1.3) %>%
                             select(OXYGEN.M) %>% pull()),
         temp.bottom = replace(temp.bottom, 
                           SITE_ID == "NLA17_MI-10073" & VISIT_NO == 1,
                           temp.o2.2 %>% filter(SITE_ID=="NLA17_MI-10073",
                                              VISIT_NO == 1,
                                              DEPTH == 1.3) %>%
                             select(TEMPERATURE.M) %>% pull()))
```

No surface temp at NLA17_NE-10060, visit 1 because no measurement taken at shallowest depth (0m). Replace with next shallowest measurement (0.1m)

```{r}
# look at raw data
temp.o2.2 %>% filter(SITE_ID == "NLA17_NE-10060", VISIT_NO == 1)

o2.s.b <- o2.s.b %>%
  mutate(temp.surf = replace(temp.surf, 
                           SITE_ID == "NLA17_NE-10060" & VISIT_NO == 1,
                           temp.o2.2 %>% filter(SITE_ID=="NLA17_NE-10060",
                                              VISIT_NO == 1,
                                              DEPTH == 0.1) %>%
                             select(TEMPERATURE.M) %>% pull()))
```

No bottom temp or O2 at NLA17_NH-10063, visit 1 because no measurement taken at greatest depth (8m). Replace with next deepest measurement (7.5m)

```{r}
# look at raw data
temp.o2.2 %>% filter(SITE_ID == "NLA17_NH-10063", VISIT_NO == 1)

o2.s.b <- o2.s.b %>%
  mutate(temp.bottom = replace(temp.bottom, 
                           SITE_ID == "NLA17_NH-10063" & VISIT_NO == 1,
                           temp.o2.2 %>% filter(SITE_ID=="NLA17_NH-10063",
                                              VISIT_NO == 1,
                                              DEPTH == 7.5) %>%
                             select(TEMPERATURE.M) %>% pull()),
         o2.bottom = replace(o2.bottom, 
                           SITE_ID == "NLA17_NH-10063" & VISIT_NO == 1,
                           temp.o2.2 %>% filter(SITE_ID=="NLA17_NH-10063",
                                              VISIT_NO == 1,
                                              DEPTH == 7.5) %>%
                             select(OXYGEN.M) %>% pull()))
```

No surface temp or O2 at NLA17_OR-10003, visit 1 because no measurement taken at shallowest depth (0m). Replace with next shallowest measurement (0.1m)

```{r}
# look at raw data
temp.o2.2 %>% filter(SITE_ID == "NLA17_OR-10003", VISIT_NO == 1)

o2.s.b <- o2.s.b %>%
  mutate(temp.surf = replace(temp.surf, 
                           SITE_ID == "NLA17_OR-10003" & VISIT_NO == 1,
                           temp.o2.2 %>% filter(SITE_ID=="NLA17_OR-10003",
                                              VISIT_NO == 1,
                                              DEPTH == 0.1) %>%
                             select(TEMPERATURE.M) %>% pull()),
         o2.surf = replace(o2.surf, 
                           SITE_ID == "NLA17_OR-10003" & VISIT_NO == 1,
                           temp.o2.2 %>% filter(SITE_ID=="NLA17_OR-10003",
                                              VISIT_NO == 1,
                                              DEPTH == 0.1) %>%
                             select(OXYGEN.M) %>% pull()))
```

No bottom temp or O2 at NLA17_SD-10057, visit 1 because no measurement taken at deepest depth (4m). Replace with next deepest measurement (3m)

```{r}
# look at raw data
temp.o2.2 %>% filter(SITE_ID == "NLA17_SD-10057", VISIT_NO == 1)

o2.s.b <- o2.s.b %>%
  mutate(temp.bottom = replace(temp.bottom, 
                           SITE_ID == "NLA17_SD-10057" & VISIT_NO == 1,
                           temp.o2.2 %>% filter(SITE_ID=="NLA17_SD-10057",
                                              VISIT_NO == 1,
                                              DEPTH == 3) %>%
                             select(TEMPERATURE.M) %>% pull()),
         o2.bottom = replace(o2.bottom, 
                           SITE_ID == "NLA17_SD-10057" & VISIT_NO == 1,
                           temp.o2.2 %>% filter(SITE_ID=="NLA17_SD-10057",
                                              VISIT_NO == 1,
                                              DEPTH == 3) %>%
                             select(OXYGEN.M) %>% pull()))
```

Only remaining missing values are observations where field crews noted sonde malfunctions.

@@

```{r}
# are all SITE_ID values in o2 and temp data also present in dg?
o2.s.b %>% filter(!(SITE_ID %in% dg$site.id)) %>% # no, OR and AZ-10007
print(n= Inf)

@@

# dim(o2.s.b) # 1209, 6
# dim(dg) # 2369, 375

# Left join: retain all dg values, exclude non-matching bf values (i.e. OR and AZ)
dg <- left_join(dg, o2.s.b, by = c("site.id" = "SITE_ID", "visit.no" = "VISIT_NO"))  
# dim(dg) # 2369, 379, good!
```

>>>>

---
title: "Data Munging"
authors: "J. Beaulieu, R. Martin, and M. McManus"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    fig_caption: yes
    depth: 2
    number_sections: true
editor_options: 
  chunk_output_type: console
---
```{r results='hide', message=FALSE, warning=FALSE, echo=FALSE}
# load libraries
library(plyr) # for ddply, load before tidyverse
library(tidyverse) # dplyr, ggplot
library(forcats) # fct_explicit_na()
library(rLakeAnalyzer) # for buoyancy frequency
library(zoo) # interpolate NA temp/DO

@@

## Manipulate
Implement unit conversion and create source/sink column:
``` {r}
dg <- dg %>%
  # unit conversion
  mutate(dissolved.ch4 = dissolved.ch4 * 1000000, # mol/L -> umol/L
         dissolved.co2 = dissolved.co2 * 1000000, # mol/L -> umol/L,
         dissolved.n2o = dissolved.n2o * 1000000000, # mol/L -> nmol/L
         # add source/sink column
         co2.src.snk = ifelse(co2.sat.ratio > 1, "source", "sink"),
         ch4.src.snk = ifelse(ch4.sat.ratio > 1, "source", "sink"),
         n2o.src.snk = ifelse(n2o.sat.ratio > 1, "source", "sink"))
```

@@

# Inspect results of join.  Looks good.  
# Only one site missing all temperature and DO data. (NLA17_CA-10082)
# One site missing all DO data (NLA17_ID-10084)
temp.o21 %>% filter(SITE_ID %in% missing$SITE_ID) %>%
  select(SITE_ID, VISIT_NO, DEPTH, TEMPERATURE, OXYGEN) %>% 
  arrange(SITE_ID, VISIT_NO, DEPTH, TEMPERATURE, OXYGEN)

# NLA17_ID-10001 visit 1 only has temp/DO measured at surface depth.  Notes indicate probe malfunction.
# Second site visit shows similar surface DO/temp and no thermal stratification.  Replace visit 1 with visit 2
# data
temp.o21[temp.o21$SITE_ID == "NLA17_ID-10001" & temp.o21$"VISIT_NO" == 1, c("OXYGEN", "TEMPERATURE")] = 
  temp.o21[temp.o21$SITE_ID == "NLA17_ID-10001" & temp.o21$"VISIT_NO" == 2, c("OXYGEN", "TEMPERATURE", "DEPTH")] %>%
  group_by(DEPTH) %>% # Need to aggregate duplicate measurements at depth = 0
  summarise(OXYGEN = mean(OXYGEN), TEMPERATURE = mean(TEMPERATURE)) %>%  # mean by depth
  select(-DEPTH) # remove depth

@@

`r temp.o2.2 %>% filter(is.na(DEPTH)) %>% nrow()` observations report NA for DEPTH, DO, and TEMPERATURE.  These will be removed.

``` {r}
# Missing depth values
# All also have NA for DO and Temp.  Remove from dataset
temp.o2.2 %>% 
  filter(is.na(DEPTH))

dim(temp.o2.2) # 10133, 5
temp.o2.2 <- temp.o2.2 %>% filter(!is.na(DEPTH))
dim(temp.o2.2) # 10085, 5
```

`r temp.o2.2 %>% filter(is.na(TEMPERATURE.M) & is.na(OXYGEN.M)) %>% nrow()` observations report NA for temperature and DO .  These are always at the shallowest (0m) or deepest depths.  When at shallowest, there is a 0.1m measurement.  When at deepest, I'm guessing no measurements were made at that depth.  Remove these observations.

``` {r}
# NA for temp and DO
fewMissing <- temp.o2.2 %>% 
  filter(is.na(TEMPERATURE.M) & is.na(OXYGEN.M)) %>%
  select(SITE_ID, VISIT_NO) %>%
  distinct(SITE_ID, VISIT_NO)

temp.o2.2 %>% filter(paste0(SITE_ID, VISIT_NO) %in% paste0(fewMissing$SITE_ID, fewMissing$VISIT_NO))

dim(temp.o2.2) # 10085, 5
temp.o2.2 <- temp.o2.2 %>% filter(!(is.na(TEMPERATURE.M) & is.na(OXYGEN.M)))
dim(temp.o2.2) # 10081, 5
```

`r temp.o2.2 %>% filter(is.na(TEMPERATURE.M) | is.na(OXYGEN.M)) %>% nrow()` observations report NA for depth or temperature.  If these are fairly random within the depth profile, they can be imputed.  No DO data for NLA17_ID-10002 (see above).  The remaining cases are fairly random.  Will impute via linear interpolation.

```{r}
fewMissing <- temp.o2.2 %>% filter(is.na(TEMPERATURE.M) | is.na(OXYGEN.M)) %>%
  select(SITE_ID, VISIT_NO) %>%
  distinct(SITE_ID, VISIT_NO)

temp.o2.2 %>% filter(paste0(SITE_ID, VISIT_NO) %in% paste0(fewMissing$SITE_ID, fewMissing$VISIT_NO))

temp.o2.2 <- temp.o2.2 %>% 
  group_by(SITE_ID, VISIT_NO) %>%
  mutate(OXYGEN.M = na.approx(OXYGEN.M, DEPTH, na.rm = FALSE, rule = 2), #rule = 2 assigns value closest to boundary 
         TEMPERATURE.M = na.approx(TEMPERATURE.M, DEPTH, na.rm = FALSE, rule = 2)) %>%
  ungroup()

dim(temp.o2.2) #10081, 5
```

@@

bf <- temp.o2.2 %>%
  filter(is.finite(TEMPERATURE.M)) %>% # This completely removes the CA site with only NA for temp data.
  group_split(SITE_ID, VISIT_NO) %>% # split into list of df
  lapply(function(x) {
    data.frame(BF = buoyancy.freq(wtr = x$TEMPERATURE.M, depths = x$DEPTH), # calculate buoyancy frequency
               SITE_ID = unique(x$SITE_ID), # add site ID
               VISIT_NO = unique(x$VISIT_NO), # add visit no
               stringsAsFactors = FALSE)}) %>% 
  lapply(function(x) x %>% mutate(DEPTHS.BF = attr(BF, "depths"))) %>% # depths stored as attribute, assign to column
  do.call("rbind", .) # collapse to df

# there are two non-finite bf values, lets have a look
bf.na <- bf %>% filter(!is.finite(BF)) %>%
  select(SITE_ID, VISIT_NO) %>%
  mutate(SITE.VISIT = paste0(SITE_ID, VISIT_NO))

@@

# These site have a single temp measurement made at shallow depths (<=1m, NLA17_MD-10004, NLA17_ND-10189).
# Presumably these are shallow sites, therefore unlikely to be stratified.  Assign a bf very close to 0 (see below).
bf %>% filter(paste0(SITE_ID, VISIT_NO) %in% bf.na$SITE.VISIT) 
temp.o2.2 %>% filter(paste0(SITE_ID, VISIT_NO) %in% bf.na$SITE.VISIT)

@@

`r bf.max %>% summarize(n.na = sum(!is.finite(bf.max$MAX.BF))) %>% pull()` of the site have a non-finite value for maximum buoyancy frequency value.  These are the two shallow sites with only a single temperature measurement.  Assign these the minimum bf.max value.
``` {r}
bf.max %>% filter(!is.finite(MAX.BF))

min.max.bf <- bf.max %>% 
  filter(is.finite(MAX.BF), MAX.BF > 0) %>%
  summarize(min.max.bf = min(MAX.BF))

@@

## Calculate thermocline depth

Dissolved gas concentrations collected near the water surface may be more related to mean conditions in the upper mixed layer than conditions near the air-water interface (e.g. depth = 0.1).  Below we calculate depth to thermocline.
The thermocline data contains information from the extra Oregon sites (see above) and the AZ site where the gas samples were unlabeled and not analyzed.  This is as expected.  No thermocline data for the California site (CA-10082) where the sonde was reported missing/malfunctioning.

``` {r}
therm <- temp.o2.2 %>%
  group_by(SITE_ID, VISIT_NO) %>%
  summarise(thermD = thermo.depth(TEMPERATURE.M, DEPTH))

summary(therm$thermD) # 395 sites with no evidence of stratification
dim(therm) #1209, 3 collapsed temp.o2.2 into one observation per site x visit

# Merge therm depth into temp.o2.2 for calcs below.  A single therm depth repeated for each row in unique site x visit
dim(temp.o2.2) # 10081, 5
dim(therm) #1209, 3
temp.o2.2 <- left_join(temp.o2.2, therm)
dim(temp.o2.2) # 10081, 6
# ggplot(dg, aes(thermD, MAX.BF)) + geom_point() + ylim(0,0.04)
```

## Surface mixed layer DO and temperature

Average all values between surface and thermocline.  The surface mixed layer data contains information from the extra Oregon sites (see above) and the AZ site where the gas samples were unlabeled and not analyzed.  This is as expected.  No mixed layer data for the California site (CA-10082) where the sonde was reported missing/malfunctioning.

```{r}
mixed.layer.do.temp <- temp.o2.2 %>%
  group_split(SITE_ID, VISIT_NO) %>% # split into list of dataframes
  lapply(function(x) { # apply to each element of list
    data.i <- if(any(is.na(x$thermD))) { # if not stratified (thermD is NA or NaN)
      x # then all data
    } else { 
      filter(x, DEPTH <= thermD) # if stratified, then only data at depths <= thermD
    }
    OXYGEN.EPI <- mean(data.i$OXYGEN.M, na.rm = TRUE) # calculate mean DO
    TEMP.EPI <- mean(data.i$TEMPERATURE.M, na.rm = TRUE) # calculate mean Temp
    # dump into dataframe
    data.frame(SITE_ID = unique(x$SITE_ID), # extract site id
               VISIT_NO = unique(x$VISIT_NO), # extract visit no
               OXYGEN.EPI = OXYGEN.EPI,
               TEMP.EPI = TEMP.EPI,
               thermD = unique(x$thermD), # identical values within each df, extract unique
               stringsAsFactors = FALSE)
  }
  ) %>% bind_rows() # collapse to df.  dplyr implementation of do.call(rbind, list)

# are all SITE_ID values in mixed.layer.do.temp also present in dg?
mixed.layer.do.temp %>% filter(!(SITE_ID %in% dg$site.id)) %>% # no, OR and AZ-10007, as expected
distinct(SITE_ID, VISIT_NO)

# are all site.id values in dg also present in mixed.layer.do.temp?
# Just CA site, as expected.
dg %>% filter(!(site.id %in% mixed.layer.do.temp$SITE_ID)) %>% # just CA site, as expected
  select(site.id, visit.no)

# merge with dg
dim(dg) # 2369, 375
dim(mixed.layer.do.temp) #1209, 5
dg <- left_join(dg, mixed.layer.do.temp, by = c("site.id" = "SITE_ID", "visit.no" = "VISIT_NO"))
dim(dg) # 2369, 378
```

@@

These calculations produced 1 site visit with at least one NA for surface/bottom temp/DO.  This is NL17_ID-10084 with no DO data (see above)

@@

```{r}
# are all SITE_ID values in o2 and temp data also present in dg?
o2.s.b %>% filter(!(SITE_ID %in% dg$site.id)) %>% # no, OR and AZ-10007, as expected
print(n= Inf)

@@

# dim(o2.s.b) # 1209, 6
# dim(dg) # 2369, 378

# Left join: retain all dg values, exclude non-matching bf values (i.e. OR and AZ)
dg <- left_join(dg, o2.s.b, by = c("site.id" = "SITE_ID", "visit.no" = "VISIT_NO"))  
# dim(dg) # 2369, 382, good!
```
